---
title: "Hand-on R class"
output: html_document
---

#### Your instructors
This file contains all the code from the NICAR 2016 hands-on R workshop. It was created by your instructors, 

Coulter Jones, cjones@medpagetoday.com

Amelia McNamara, amcnamara@smith.edu



#### Some background info

R has a bit of a learning curve but once you get used to it, you’re off to the races. Keep in mind you will get a ton of errors while you’re learning R — in fact, you’ll still deal with errors even if you’re a master user! Google is your friend — copy and paste the error into Google and see what others have written.

> PRO TIP: R is VERY VERY VERY case-sensitive. A misplaced capital letter or double-quote where there should be a single-quote might just be the culprit to your error.

And don’t forget — save EVERYTHING! RStudio allows you to save your data, your scripts, your history, your entire RStudio workspace — save it all and save often.

#### RStudio 
RStudio is a great interface to the R language. The interface has 4 panes. 

You’ll do your coding in the **Console** (lower left, looks like Terminal/command line). 

The **Environment and History** (upper right) panes keep track of everything you have loaded into your environment (mostly, data) and all the R commands you have ever typed. This is great because you can search through your past code to find commands you may have forgotten.  

The **Source** pane is on the upper left and displays your data tables, as well as any documents you are reading or editing (maybe, you're looking at this document there!). 

The **Files/Plots/Packages/Help/View** pane on the lower right does a bunch of stuff — holds files, displays the graphs you’ve made, shows the packages you have installed, and more. Perhaps the most important tab of that lower right pane is Help. R has two help functions. Forget what an average or mean is? Try this.

```{r, eval=FALSE}
?mean
```
or
```{r, eval=FALSE}
help(mean)
```

#### Looking for more help? 

- The main R site: [http://cran.r-project.org/]( http://cran.r-project.org/) This has everything from guides to R updates. It also has a comprehensive list of packages and lots of great tips on how to use R. 
- Official R Language Definition: [http://cran.r-project.org/doc/manuals/r-release/R-lang.html](http://cran.r-project.org/doc/manuals/r-release/R-lang.html)
- An In-Depth Introduction to R: [http://cran.r-project.org/doc/contrib/usingR.pdf](http://cran.r-project.org/doc/contrib/usingR.pdf)
- Using R for Intro Stats: [http://cran.r-project.org/doc/contrib/Verzani-SimpleR.pdf](http://cran.r-project.org/doc/contrib/Verzani-SimpleR.pdf)
- Quick R: [http://www.statmethods.net/](http://www.statmethods.net/)
- R Documentation Search: [http://www.rdocumentation.org/](http://www.rdocumentation.org/)
- R-bloggers: find out awesome things people are doing with R [http://www.r-bloggers.com/](http://www.r-bloggers.com/)
- Coursera Course on R: [https://www.coursera.org/course/rprog](https://www.coursera.org/course/rprog)
- Princeton Guide to R: [http://libguides.princeton.edu/dss/R](http://libguides.princeton.edu/dss/R)
- UCLA Guide to R: [http://www.ats.ucla.edu/stat/r/sk/books_usingr.htm](http://www.ats.ucla.edu/stat/r/sk/books_usingr.htm)
- Stack overflow: [http://stackoverflow.com/](http://stackoverflow.com/)


#### Packages and syntax 

R has "base" functionality that is built into the language, but most of the good stuff comes from user-written "packages" that extend the language. We'll be using the `mosaic` and `ggplot2` packages in this class, among others. Because R is open source (and was written by statisticians...) it has some strange features compared to other programming languages. The most obvious is that its syntax is not consistent. Almost any task you want to do, there are three ways to accomplish. And packages can contribute to this nonsense by providing their own syntax. We'll try to point to places where the code we are showing you is just one of many ways to do something, but know that when you start googling questions you may see things that don't look familiar. Don't be afraid!

To use a package, you have to `require()` it. Lets start by requiring some packages:
```{r, message=FALSE, warning=FALSE}
require(mosaic)
require(ggplot2)
```


#### Data!
Later in this class, we will show you how to load in external data, but for now we are going to begin by playing with some data that comes with R. This will allow us to start playing quickly.

We'll start with some data about midwest demographics. To get started, you can read the data in with the `data()` command, and see the first few rows with `head()`.

```{r}
data(midwest)
head(midwest)
```
For a spreadsheet-like experience of your data, just click on its name in your Environment tab to see a preview of the data in the Source pane. Note that you can only look, you can't "touch." R is a programming language, so if you see values you want to have changed, or you need to add another variable based on an existing one you have to do so programmatically. 

If you want to read about all the variables that are included in the dataset, you can run
```{r}
?midwest
```
or
```{r}
help(midwest)
```

A few more handy functions that have to do with datasets are `str()`, which tells you about the **str**ucture of your data,
```{r}
str(midwest)
```
and `summary()`, which provides a summary.
```{r}
summary(midwest)
```

You can also get the list of variable names using the `names()` function,
```{r}
names(midwest)
```

#### Plotting in R
Now that we have data, we can begin doing Exploratory Data Analysis, making some plots to give us a sense of the relationships in the data.

Here's how we can create a scatterplot:
```{r}
xyplot(popwhite~popblack, data=midwest)
```

#### Summary statistics
Another part of exploratory data analysis is finding summary statistics. These work much the same way as plots. 

```{r}
mean(~popblack, data=midwest)
```

Other than `mean()` you can try `median()` `mode()` `min()` `max()` `sd()` and many more. Or, for a best-of function, try `favstats()`

```{r}
favstats(~popblack, data=midwest)
```

If you have two variables and want to see their relationship, you can feed most of these functions several arguments. For example,
```{r}
mean(popblack~inmetro, data=midwest)
```

1. Do metro areas have larger percentages of whites than rural areas? How do you know?

#### Sorting and ordering
Maybe we're working on a story about the county in the midwest that has the largest percentage of black people. We can use the `arrange()` function to see this.
```{r}
midwest %>% arrange(percblack)
```

But, this is sorting in increasing order, and we want in decreasing order. We could look in the documentation for `arrange()` to see how to change this.
```{r}
midwest %>% arrange(desc(percblack))
```

2. What county has the largest American Indian population (in absolute numbers)?

