---
title: "Hands-on R class"
output: 
  html_document:
    css: lab.css
---
```{r, message=FALSE, warning=FALSE, echo=FALSE}
require(knitr)
opts_chunk$set(eval=FALSE)
```



## Your instructors
This file contains all the code from the NICAR 2016 hands-on R workshop. It was created by your instructors, 

Coulter Jones, coulter.jones@wsj.com

Amelia McNamara, amcnamara@smith.edu


## Some background info

R has a bit of a learning curve but once you get used to it, you’re off to the races. Keep in mind you will get a ton of errors while you’re learning R — in fact, you’ll still deal with errors even if you’re a master user! Google is your friend — copy and paste the error into Google and see what others have written.

> PRO TIP: R is VERY VERY VERY case-sensitive. A misplaced capital letter or double-quote where there should be a single-quote might just be the culprit to your error.

And don’t forget — save EVERYTHING! RStudio allows you to save your data, your scripts, your history, your entire RStudio workspace — save it all and save often.

## RStudio 
RStudio is a great interface to the R language. The interface has 4 panes. 

You’ll do your coding in the **Console** (lower left, looks like Terminal/command line). 

The **Environment and History** (upper right) panes keep track of everything you have loaded into your environment (mostly, data) and all the R commands you have ever typed. This is great because you can search through your past code to find commands you may have forgotten.  

The **Source** pane is on the upper left and displays your data tables, as well as any documents you are reading or editing (maybe, you're looking at this document there!). 

The **Files/Plots/Packages/Help/View** pane on the lower right does a bunch of stuff — holds files, displays the graphs you’ve made, shows the packages you have installed, and more. Perhaps the most important tab of that lower right pane is Help. R has two help functions. Forget what an average or mean is? Try this.

```{r, eval=FALSE}
?mean
```
or
```{r, eval=FALSE}
help(mean)
```

## Looking for more help? 

#### Books
- R Cookbook. Both a website, [http://www.cookbook-r.com/](http://www.cookbook-r.com/), and a book, [http://smile.amazon.com/gp/product/1449316956/](http://smile.amazon.com/gp/product/1449316956/)
- A Compedium of Commands to Teach Statistics with R. Both a downloadedable pdf, [http://mosaic-web.org/go/Master-Core.pdf](http://mosaic-web.org/go/Master-Core.pdf), and a book, [http://smile.amazon.com/Compendium-Commands-Teach-Statistics/dp/0983965889](http://smile.amazon.com/Compendium-Commands-Teach-Statistics/dp/0983965889)
- Start Teaching with R. Both a downloadable pdf, [http://mosaic-web.org/go/Master-Starting.pdf](http://mosaic-web.org/go/Master-Starting.pdf) and a book, [http://smile.amazon.com/Start-Teaching-R-Randall-Pruim/dp/0983965862/r](http://smile.amazon.com/Start-Teaching-R-Randall-Pruim/dp/0983965862/r)
- Using R for Intro Stats: Downloadable pdf, [http://cran.r-project.org/doc/contrib/Verzani-SimpleR.pdf](http://cran.r-project.org/doc/contrib/Verzani-SimpleR.pdf)

#### Cheatsheets 
Do you like having something printed out and at your side at all times?

- RStudio cheatsheets: [http://www.rstudio.com/resources/cheatsheets/](http://www.rstudio.com/resources/cheatsheets/) On the RStudio interface, packages (dplyr, ggplot2), RMarkdown, etc. 
- Minimal R: [https://cran.r-project.org/web/packages/mosaic/vignettes/MinimalR.pdf](https://cran.r-project.org/web/packages/mosaic/vignettes/MinimalR.pdf)

#### Website resources
- R Cookbook. Both a website, [http://www.cookbook-r.com/](http://www.cookbook-r.com/), and a book, [http://smile.amazon.com/gp/product/1449316956/](http://smile.amazon.com/gp/product/1449316956/)
- Quick R: [http://www.statmethods.net/](http://www.statmethods.net/)
- R-bloggers: find out awesome things people are doing with R [http://www.r-bloggers.com/](http://www.r-bloggers.com/)

#### Online learning/courses
- Johns Hopkins Coursera Course on R: [https://www.coursera.org/course/rprog](https://www.coursera.org/course/rprog). Part of the Data Science specialization, [https://www.coursera.org/specializations/jhu-data-science](https://www.coursera.org/specializations/jhu-data-science). Courses are free, but specialization costs money. 
- DataCamp: [https://www.datacamp.com/](https://www.datacamp.com/) Interactive way to learn R in the browser. Free to start, pretty cheap to continue. 

#### Searching
- R Documentation Search: [http://www.rdocumentation.org/](http://www.rdocumentation.org/)
- Stack overflow: [http://stackoverflow.com/](http://stackoverflow.com/) You'll want to put [r] in your search box.








## Packages and syntax 

R has "base" functionality that is built into the language, but most of the good stuff comes from user-written "packages" that extend the language. We'll be using the `mosaic`, `ggplot2` and `dplyr` packages in this class, among others. Because R is open source (and was written by statisticians...) it has some strange features compared to other programming languages. The most obvious is that its syntax is not consistent. Almost any task you want to do, there are three ways to accomplish. And packages can contribute to this nonsense by providing their own syntax. We'll try to point to places where the code we are showing you is just one of many ways to do something, but know that when you start googling questions you may see things that don't look familiar. Don't be afraid!

To use a package, you have to `require()` it. Lets start by requiring some packages:
```{r, message=FALSE, warning=FALSE}
require(mosaic)
require(ggplot2)
```

## Data!
Later in this class, we will show you how to load in external data, but for now we are going to begin by playing with some data that comes with R. This will allow us to start playing quickly.

We'll start with some data about midwest demographics. To get started, you can read the data in with the `data()` command, and see the first few rows with `head()`.

```{r}
data(midwest)
head(midwest)
```
For a spreadsheet-like experience of your data, just click on its name in your `Environment` tab to see a preview of the data in the Source pane. Note that you can only look, you can't "touch." R is a programming language, so if you see values you want to have changed, or you need to add another variable based on an existing one you have to do so programmatically. 

If you want to read about all the variables that are included in the dataset, you can run
```{r}
?midwest
```
or
```{r}
help(midwest)
```

A few more handy functions that have to do with datasets are `str()`, which tells you about the **str**ucture of your data,
```{r}
str(midwest)
```
and `summary()`, which provides a summary.
```{r}
summary(midwest)
```

You can also get the list of variable names using the `names()` function,
```{r}
names(midwest)
```

## Changing data types
Sometimes, the way that R has loaded in data is not exactly what we would like. In that case, we might want to change the data type. For example, `inmetro` has values of 0 and 1, with 0 meaning "not in metro" and 1 meaning "in metro." R has considered this to be `integer` data, but we would rather have it as `factor` data (R's word for categorical).

```{r}
midwest <- midwest %>%
  mutate(inmetro = factor(inmetro))
```

To do this transformation, we are overwriting the old version of our `midwest` dataset with a new version. This change takes the original dataset, and then "pipes" the data (using the `%>%` operator) into a `mutate()` call. 

1. Try running `str()` on the dataset again. What happened to `inmetro`?
2. How would you change the variable `category` to be a factor?

## Plotting in R
Now that we have data, we can begin doing Exploratory Data Analysis, making some plots to give us a sense of the relationships in the data.

Here's how we can create a scatterplot:
```{r}
xyplot(popwhite~popblack, data=midwest)
```

To see the distribution of one variable, try a histogram or a density plot
```{r}
histogram(~percprof, data=midwest)
densityplot(~percprof, data=midwest)
bwplot(~percprof, data=midwest)
```

Maybe you want to compare a few different groups. You could make two plots side by side,
```{r}
histogram(~percprof|inmetro, data=midwest)
bwplot(~percprof|inmetro, data=midwest)
```
You could put two plots on the same page,
```{r}
densityplot(~percprof, groups=inmetro, data=midwest, auto.key=TRUE)
bwplot(inmetro~percprof, data=midwest)
```
Or you could color points by another variable, 
```{r}
xyplot(poptotal~area, groups=inmetro, data=midwest)
```


## Summary statistics
Another part of exploratory data analysis is finding summary statistics. These work much the same way as plots. 

```{r}
mean(~popblack, data=midwest)
```

Other than `mean()` you can try `median()` `mode()` `min()` `max()` `sd()` and many more. Or, for a best-of function, try `favstats()`

```{r}
favstats(~popblack, data=midwest)
```

If you have two variables and want to see their relationship, you can feed most of these functions several arguments. For example,
```{r}
mean(popblack~inmetro, data=midwest)
```

1. Do metro areas have larger percentages of whites than rural areas? How do you know?

For categorical data, `tally()` is useful,
```{r}
tally(~inmetro, data=midwest)
```
And, it can be used to tally two variables together,
```{r}
tally(inmetro~category, data=midwest)
```

R is also fond of chaining together functions, or nesting them. So, once you've `tally`-ed your data you can use those counts to make a barchart. 
```{r}
barchart(tally(~category, data=midwest))
```

## Sorting and ordering
Maybe we're working on a story about the county in the midwest that has the largest percentage of black people. We can use the `arrange()` function to see this.
```{r}
midwest %>% arrange(percblack)
```

But, this is sorting in increasing order, and we want in decreasing order. We could look in the documentation for `arrange()` to see how to change this.
```{r}
midwest %>% arrange(desc(percblack))
```

2. What county has the largest American Indian population (in absolute numbers)?

<<<<<<< HEAD
#### Second hour of panels
Make sure you have the packages loaded if you don't already.
```{r}
require(mosaic)
require(ggplot2)
require(dplyr)
```
=======
>>>>>>> e9ad16cb27cc624d9655d4e531fff4f299e48856

## Loading Data
Most of the time, you won't be working with data sets available through packages and will need to import data in to R. RStudio makes this very simple. On the **Environment** pane (upper right corner), the **Import Dataset** icon provides a step-by-step process to importing files. Following those steps will not only import the data, but print out the code in the **Console** showing how it was done.

```{r}
docs <- read.csv("~/Desktop/PartD_Providers_FORCLASS.csv", header = T)
```

To import this data set, we used the `read.csv()` function. The key part of this command is making sure that you direct the command to where the file is located on your computer. Check other functionality of `?read.csv` if you have questions.

> **Note:** The way file directories are referenced in R varies between a Mac and PC. 

> MAC file directories follow this protocol use "~/file_location"

> PC versions would follow this: "C:\file_lcocation\"

Let's see what our data looks like to make sure it imported correctly. Remember, the `str()` command tells you the structure of your data.
```{r}
str(docs)
```

The data appears to have imported correctly for reach field. We can move on to some basic summary statistics.
```{r}
summary(docs)
```

Although interesting, let's really focus in one field:*COST_SUM*. `favstats`, which we used on the other data set will show the general flow.

```{r}
favstats(~COST_SUM, data=docs)
```

#### Quick charts with ggplot
We see from our stats function that the **median (298,512)** is significantly lower than the **mean (401,605.7)**. We know that might be a sign of some outliers on he higher end, particularly if we read **Sarah's Cohen's Numbers in the Newsroom** [http://store.ire.org/products/numbers-in-the-newsroom-using-math-and-statistics-in-news-second-edition-e-version]. Let's chart our data to see the distribution.

The `ggplot` library has many charting capabilities, but `qplot()` function is the quickest way to quickly visualize your data. Let's make a histogram of *COST_SUM*.

```{r}
qplot(COST_SUM, data = docs)
```

<<<<<<< HEAD
As a function qplot basically works like this: ***qplot([x_variable], [y_variable], data = [data_frame], geom = "[geom_type]")*** (Remember, use ```?qplot()``` or ```help(qplot)``` if you need help.) The function makes assumptions on the best type of chart to use. In the above example, because you only passed one variable it assumed a histogram is what is needed. We could explictly tell it to do that. 

```{r}
qplot(COST_SUM, data = docs, geom = "histogram")
```

The histogram is revealing, a similar type of chart is ***density*** which also shows the distribution.
```{r}
qplot(CLAIM_COUNT, data = docs, geom = "density")
```

So far we've only charted one variable. Maybe we want to look at two variables in comparison. For example, How the total claims per physician (***CLAIM_COUNT***) compares to the number of of brand name claims (***BRAND_COUNT***). 

```{r}
qplot(CLAIM_COUNT, BRAND_COUNT, data = docs)
```

Maybe, we ant to look at that as line graph, instead.

```{r}
qplot(CLAIM_COUNT, BRAND_COUNT, data = docs, geom = "line")
```

That, isn't super helpful. What we really want to see is a general ***smooth*** line showing the general trend. 
=======
## Quick charts with ggplot
Charting 
>>>>>>> e9ad16cb27cc624d9655d4e531fff4f299e48856

```{r}
qplot(CLAIM_COUNT, BRAND_COUNT, data = docs, geom = "smooth")
```

Sadly, that isn't that interesting. The trend shows us the trend AND the individual plotted points to see who is above or below the general trend line. You can use the ***combine function*** ```c()``` to chart both.
```{r}
qplot(CLAIM_COUNT, BRAND_COUNT, data = docs, geom = c("point", "smooth"))
```

## Altering a data frame in R
There is plenty to work with for this data, but let's say we want to alter it slightly and calculate additional fields. If we type equations out, the **Console** wil return the results.
```{r}
2+4
```

R follows standard mathematical order of operations **(PEMDAS)**. So, this:
```{r}
2+4*3^2
```

gives you a different result than this:
```{r}
(2+4)*3^2
```

As journalists, it's pretty rare to just be type out equations like that. However, it's not uncommon to use math on our existing data to create new fields. Take the **CLAIM_COUNT** and **BRAND_COUNT** fields. We know the total number of claims each doctor has approved and how many of those claims were filled with brand name drugs. We need to divide **BRAND_COUNT** by **CLAIM_COUNT**. We can use `mutate()` to create a field in our data called **BRAND_PCT** with this ratio.

```{r}
docs <- mutate(docs, BRAND_PCT = BRAND_COUNT/CLAIM_COUNT)
```

and `head()` and `str()` to take a look at the data field we just created.
```{r}
head(docs)

str(docs)
```
While the ratio is correct, we'd prefer to have it look more like a percentage by multiplying the ratio by 100. We can simply overwrite the existing data we just created.

```{r}
docs <- mutate(docs, BRAND_PCT = (BRAND_COUNT/CLAIM_COUNT)*100)

head(docs)
```

## Filtering data with DPLYR
Now this has been interesting, but in many cases we want to pull out specific pieces of our data. What if we're only interested in doctors in Colorado. We can do this with the `filter()` function, which is part of the `dplyr` package. Remember, if this is our first time using `dplyr` this sessione we  need to `require()` first.
```{r}
require(dplyr)
```

This is how the filter function works. `filter(data, method you want to filter on)`. If we use it will simply print the results in the **Console**. We can also assign the results to a new dataframe. To find only doctors from Colorado:

```{r}
colorado_docs <- filter(docs, STATE == "CO")

View(colorado_docs)
```

You don't have to be limited by one variable. Earlier we looked at high-cost prescribers. Perhaps, we'd like to do that just for Colorado doctors. We can use the `&` to to look at all Colorado doctors who have

BOOLEANS review: In R we can use two types of Boolean characters 

**AND** represented by **&**

**OR** represented by **|**

**&** limits our filter because both criteria must be met. **|** broadens our filter. If either criteria is met the data is included.

The above code shows doctors just from Colorado who also exceeded $500,000 in costs. Below would give doctors who either are from Colorado or have a cost that exceeds $500,000 and could be from any state.
```{r}
colorado_high <- filter(docs, STATE == "CO" & COST_SUM > 500000)
```

Now we can view the new data set we created in our **Environment** or order by using `arrange()` which you learned earlier

```{r}
head(colorado_high, 20) %>% arrange(desc(COST_SUM))
```

## Grouping with dplyr
That's helpful, but what if you want to get a sense of the average salary or total number of doctors in each state. In Excel you might use a PivotTable, in SQL you'd used a "Group By"" query. In `dplyr` the syntax is very similar to SQL with the `group_by()` and `summarise()` functions.

You'll do this in two steps. First, decide which variables to put in to the `group_by` function. 
```{r}
state_group <- group_by(docs, STATE)
```

Then use what you just created to summarise our data and then view the result.
```{r}
state_docs <- summarise(state_group,
  count = n(),
  median_cost = median(COST_SUM)
)

state_docs
```

You can add as many summary statistics as you want.
```{r}
state_docs <- summarise(state_group,
  count = n(),
  median_cost = median(COST_SUM),
  total_cost = sum(COST_SUM),
  high_cost = max(COST_SUM),
  sd_cost = sd(COST_SUM),
  median_scripts = sum(CLAIM_COUNT),
  total_scripts = sum(CLAIM_COUNT)
)
```

If you decided you wanted to group by more than one variable, just add it to the function, separating each variable by a comma. To group by city and state would be: 

```{r}
state_group <- group_by(docs, CITY, STATE)
```

Then, just re-run the summarise function. 
```{r}
state_docs <- summarise(state_group,
  count = n(),
  median_cost = median(COST_SUM),
  total_cost = sum(COST_SUM),
  high_cost = max(COST_SUM),
  sd_cost = sd(COST_SUM),
  median_scripts = sum(CLAIM_COUNT),
  total_scripts = sum(CLAIM_COUNT)
)
```

#### Other Options
We've shown you several R functions, but keep in mind. The syntax for R's base package, for example, is slightly different.

***[data_frame]$[variable]***

So for example, if you want


